---
title: "Covid_Mixed_effect_analysis"
author: "Jingheng C"
date: "11/24/2021"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
## Data source
https://github.com/CSSEGISandData/COVID-19

## Introduction

In this analysis, we want to examine and quantify how the COVID19 pandemic impacted different parts of the globe in different ways. 

Source code: https://github.com/IanC544/statistical_analysis_projects


```{r cars}
library(tidyverse)
```

## Exploratory data analysis with Visualization

```{r}
data = read.csv('WHO-COVID-19-global-data.csv')
summary(data)
#There is often a lag between COVID-19 cases and deaths, so we manually 
#introduce a 7 day lag for better analysis
data = data %>% group_by(Country) %>%
  mutate(Cumulative_deaths = lag(Cumulative_deaths,n=7, default = NA))
#We want to learn about the COVID pandemic at its peak, so we only take the 
#30 days with the highest new cases count.
data = data  %>% group_by(Country) %>% 
  slice_max(order_by = Cumulative_deaths, n=30)


#Initial visualization - for clarity, I only plot the five countries with the 
#highest COVID deaths
top5 = data %>% group_by(Country) %>% 
  summarise(count = max(Cumulative_deaths)) %>% 
  slice_max(order_by = count, n=5)

data %>% filter(Country %in% top5$Country) %>% 
  ggplot(aes(x = Cumulative_cases,y = Cumulative_deaths,group = Country)) +
  geom_smooth(aes(color=Country),method = "lm")+
  geom_point(aes(color=Country))

```
From the above visualization, we can see the relationship between case count & the death count is very different from country to country, so we must control for the country-specific effect for new cases.  

## Mixed model with lme4
Linear Mixed-Effects Models from lme4 can take country-specific effects into account, this will allow us to better understand the variability in the number of death across different countries
```{r}
#standardization before fitting
data$Cumulative_cases=scale(data$Cumulative_cases)
data$Cumulative_deaths=scale(data$Cumulative_deaths)
library(lme4)
library(sjstats)#needed for icc
covid_mixed = lmer(Cumulative_deaths ~ Cumulative_cases +
                     (Cumulative_cases| Country),data)
summary(covid_mixed)
icc(covid_mixed)
```
## Conclusion
You might notice the previous provides no p-value, that is because the degree of
freedom can be hard to calculate for mixed designs.
The ICC(Intraclass-Correlation Coefficient) shows the majority of the variability in the number of death is due to the between-subject variations. This analysis suggests that the COVID impacted counties in different ways. In future modeling, we need to consider the country-specific effect.


## Model Diagnostics
To check whether the assumptions for lme model holds.
```{r}
coef(covid_mixed)$Country
plot(covid_mixed)
qqnorm(residuals(covid_mixed))
```
From the residual plot and the QQ plot, we can see that the residual is not normally distributed and has a nonconstant variance. The shape of our QQ plot also suggests our data is over-dispersed. Thus making the output of our lme model unreliable.


## Known biases
Due to over-dispersion, the estimated coefficients are likely to be heavily biased, but the conclusion should still hold since the source of the problem originated from the fact that the data from different countries are vastly different from one to another.
```{r}
sessionInfo()
```

 