---
title: "Analysis of NYPD Shooting Incident Data (Historic)"
author: "Jingheng C"
date: "11/17/2021"
output: pdf_document
---
A quick analysis of the NYPD Shooting Incident Data. We want to see whether 
there is a temporal trend with the incident count.

Data Publisher: data.cityofnewyork.us
Data Maintainer: NYC OpenData

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Importing data and libraries

```{r imports}
library(tidyverse)
library(lubridate)

NYPD_Shooting_data = read.csv("NYPD_Shooting_Incident_Data__Historic_.csv")

summary(NYPD_Shooting_data)
```
## Data Cleaning & Transformation
```{r Cleaning & Transformation}

# Extract year&month from OCCUR_DATE because we need to group the incident 
# by month and year for this analysis
NYPD_Shooting_data = NYPD_Shooting_data %>%
  mutate(OCCUR_DATE_Month_Year = format(mdy(OCCUR_DATE),"%Y-%m"))

# Aggregation to get the monthly incident count 

Monthly_incident_count = NYPD_Shooting_data %>% group_by(OCCUR_DATE_Month_Year) %>% 
  summarise(incident_count = n())

Monthly_incident_count$date = ym(Monthly_incident_count$OCCUR_DATE_Month_Year)

summary(Monthly_incident_count)

```




## Data Visualization

We can easily spot a huge spike in incident count around mid-2020, this spike is
likely caused by the COVID-19 recession. There seems to be a downward trend, 
further quantitative analysis is needed to conclude whether there is a trend.   


```{r Data Exploration}

timeseries_plot = ggplot(Monthly_incident_count,aes(date,incident_count))+
  geom_line()+
  xlab('Incident Count')
timeseries_plot

```


## Regression analysis of the trend
We will use the index of row to acts as a variable of the linear trend
```{r}
# Adding index column 
Monthly_incident_count$Trend = seq.int(nrow(Monthly_incident_count))

# Linear Regression
model = lm(incident_count ~ Trend,data = Monthly_incident_count)
summary(model)

```

p = 2.3e-15 < 0.05
We conclude that there is significant evidence to suggest the presence of 
a linear trend in the incident count.


# Adding more independent variable
In the previous regression analysis, R-squared is only 0.29.
Around 70% of the variance is unexplained by our model.
Seasonality seems to be present in the data,

## Data Visualization
Decompose the data to get a better visual representation of the seasonality.


```{r summary}
# Convert the data into time series
ts = ts(Monthly_incident_count$incident_count,frequency=12)

# Time series decomposition to visualize the seasonality 

ts = decompose(ts)

# Visualize the decomposed time series

plot(ts)

```





## Regression Analysis


```{r}
# Adding variable for Month
Monthly_incident_count$Month = as.factor(month(Monthly_incident_count$date))


model_se = lm(incident_count ~ Trend + Month ,data = Monthly_incident_count)

summary(model_se)

```
The p value of many seasonal dummy variables is less than 0.05.
We can conclude that seasonality is present in the data. 
By adding months to our model, Adjusted R-squared increased to 0.5858. 
A majority of the variance is now explained by our model


# Conclusion & Biases
The monthly shooting incidents count has a decreasing trend over time. Seasonal patterns can also be observed, the number of shooting incidents is significantly higher from May to October compared to other months. 

One potential source of biases is with the data collection process, the data was collected and published by the government of New York City. There could be a political incentive for the government to publish data that suggest the number of shooting incidents is decreasing over the years. Ideally a third party NGO should verify the data is authenticate 
```{r}
sessionInfo()
```
